Global search algorithm with local tuning for solving mixed-integer global optimization problems \thanks{This study was supported by the .}

### 2021-04-10 2021/MOTOR_BarkalovLebedev/MOTOR2021.tex
MOTOR правкти

Choosing the optimal decisions (decision making) in situations with many different alternatives is a problem that occurs in almost every domain of human activity. In many cases, decision making problems can be viewed as optimization problems of various kinds such as convex programming, discrete optimization, nonlinear programming, etc. In more complex situations, the objective functions that determine the effectiveness of decisions can be multiextremal, and decision making will require solving global optimization problems. In the most general case, however, the effectiveness of decisions may be determined by several objective criteria, which necessitates solving multi-objective optimization (MOO) problems. Thus, MOO problems are the most common decision making statements that one has to solve in many scientific and technical applications. A large number of approaches and methods have been developed to solve such problems and they have been used to solve many decision making problems in various fields of practical applications -- see e.g. \cite{c1,c2,c3,c4,c5,c6,c7,c8,c9}.

@@

It is also important to note that by solving simultaneously a large number of scalar optimization problems thus generated it is possible to significantly decrease computational complexity of each separate problem. This effect is achived due to the fact that all such scalar problems  are based on the same MOO problem and, consequently, all computed values of the efficiency criteria of the MOO problem can be reduced to values of any scalar problem being solved simultaneously without any time consuming calculations. In such cases, all the search information obtained in solving any single scalar problem can be used for solving all other scalar problems of the same set.

This paper presents the results of the authors continuing research aimed to develop efficient methods for solving MOO problems \cite{c35,c36,c37,c44}. In \cite{c35}, some methods for solving global optimization problems are considered, in \cite{c36},  a general decision-making model based on MOO problems is proposed, and in \cite{c37}, a theoretical analysis is presented regarding the effectiveness of the use of search information when solving MOO problems. In \cite{c44}, we consider MOO problems in which the efficiency criteria can be ordered by importance (lexicographic multi-objective optimization). In this case, MOO problems are reduced to global optimization problems with non-convex constraints for which constrained optimization methods should be applied. In this paper, we propose new computational schemes for using search information that significantly reduce the amount of calculations required to solve MOO problems.

@@

The reference decision $y^0 \in D$ in (\ref{eq:05}) can be known \textit{a priori} or determined on a basis of some known prototype. In many cases, an abstract ideal decision $y^0 \in D$ is used as the reference decision $y^*\in D$ in which the objective functions $f_i$, $1 \leq i \leq s$, have minimal possible values, i.e.
\begin{equation}
\label{eq:07}
f_i^* =f_i (y^*)=\min_{y \in D} f_i (y), 1 \leq i \leq s.
\end{equation}

@@

Using the dimensionality reduction results in one more additional advantage of the proposed approach: many well-known one-dimensional global search algorithms (possibly, after some additional generalisation) can be used to solve the initial multidimensional MOO problem from (\ref{eq:01}) \cite{c22,c23,c24,c25,c26,c27,c28,c29}. At the same time, it should be noted that most works where dimensionality reduction is a key feature for solving multiextremal problems like (\ref{eq:04}) rely on the information-statistical theory of global search \cite{c15}. This theory has provided the a basis for developing a large number of efficient methods for multiextremal optimization \cite{c28,c29,c30,c31,c32,c33,c34,c35,c36,c37}.

@@

Convergence conditions of the algorithms formulated in the form (\ref{eq:16})--(\ref{eq:18}) depend on the properties of the interval characteristics used. One of sufficient conditions for convergence of algorithms is, for example, the requirement that the characteristic of the interval containing the global minimum point should take on the maximum value at step 1 of the scheme (\ref{eq:16})--(\ref{eq:18}) during the global search iterations. This condition is satisfied, for example, for the multidimensional generalized algorithms proposed in \cite{c22,c23} when the H{\"o}lder constant from (\ref{eq:10}) is specified exactly. For GSA, a sufficient condition for convergence is the relation 
\begin{equation}
\label{eq:20}
m \geq 2^{3-1⁄N} L\sqrt{N+3},
\end{equation}
which must be fulfilled starting from some iteration $k>1$ of the global search ($L$ is the Lipschitz constant from (\ref{eq:08})) \cite{c15}. Moreover, if condition (\ref{eq:20}) is satisfied, only the points of the global minimum of the function $\varphi(y)$ from (\ref{eq:04}) will be the limit points of the trial sequence $\{y^k=y(x^k)\}$ generated by the GSA algorithm.

@@

As mentioned earlier, in the process of solving the MOO problem, it may be necessary to find several different efficient decisions due to possible changes in the optimality requirements. Obtaining different efficient decisions in the proposed approach is ensured by choosing different convolution coefficients (importance indicators) for the objective functions $f_i(y)$, $1 \leq i \leq s$, which results in obtaining different  scalar multiextremal functions $F(\lambda, y)$ from (\ref{eq:04}). These functions $F(\lambda, y)$ can be optimized sequentially. This determines a multi-stage scheme for solving the MOO problem, when at each stage DM specifies the necessary scalarization coefficients, then the resulting scalar optimization problem is solved, after which DM analyzes the efficient decisions found.

@@

As shown in Section \ref{sec:03}, the information-statistical multiextremal optimization algorithms used in the proposed approach determine the points of consecutive iterations of the global search taking into account the search information
\begin{equation}
\label{eq:22}
A_k=\{(x_i,z_i,f_i )^T:1 \leq i \leq k\}
\end{equation}
obtained in the calculations (see (\ref{eq:14})--(\ref{eq:17})). In (\ref{eq:22}) $x_i$, $1 \leq i \leq k$,  are the reduced points of performed global search iterations ordered in ascending order of coordinates,  $z_i$, $f_i$, $1 \leq i \leq k$, are values of the scalar function $F(\lambda, y)$, from (\ref{eq:04}) and objective functions $f_i(y)$, $1 \leq i \leq s$, from (\ref{eq:01}) of the current optimization problem to be solved at the points $x_i=y(x_i)$, $1 \leq i \leq k$. By using search information $A_k$ from (\ref{eq:22}) when choosing next search iterations it is possible to solve global optimization problems more efficiently and to provide convergence of algorithms only to the global minima of multiextremal functions being minimized.

@@

For MGSA 100 problems (\ref{eq:04}) with different values of convolution coefficients $\lambda$ uniformly distributed in $\Lambda$ were solved. The results of the experiments performed are presented in Table \ref{tab:1}.

@@

The results of numerical experiments are presented in Table \ref{tab:2} and Fig. \ref{fig:1}. The Table shows in the first column the number of functions $\Phi_p (y)$ from (\ref{eq:21}) being optimized.  The columns named Iters contain the average number of iterations performed by the algorithm to solve the MOO problem. The columns named HV and DU contain the values of the HV and DU indicators. The results of the numerical experiments are divided with respect to the three algorithms used: GSA is described in Section \ref{sec:03}, MGSA is discussed in Section \ref{subsec:32}, SGSA is presented in Section \ref{subsec:43}.

@@

The experimental results show that by reusing search information when optimizing 25 functions from the set $\Phi_p (y)$ the total computational iterations can be reduced  by more than 9.5 times without resorting to any additional computational resources. At the same time, the MGSA algorithm performs the least number of iterations (the number of calculations of objective function values), while the SGSA algorithm provides a better approximation of the Pareto domain both in terms of approximation completeness and coverage uniformity. It is also noteworthy that the average number of iterations to optimize one function from the set $\Phi_p (y)$ decreases by more than 4 times as the number of functions in the set $\Phi_p (y)$ increases -- see Table \ref{tab:3}.

@@

In the third series of numerical experiments, 10 bi-criteria five-dimensional MOO problems were solved, i.e. $N = 5$, $s = 2$. The objective functions of the MOO problems were determined, as before, using the GKLS generator \cite{c43}. When solving the problems, the accuracy $\varepsilon=0.05$ and the reliability $r=5.6$ were used. The results of numerical experiments are presented in Table \ref{tab:4} and Fig.\ref{fig:2}.

>>>>

Choosing the optimal decisions (decision making) in situations with many different alternatives is a problem that occurs in almost every domain of human activity. In many cases, decision making problems can be viewed as optimization problems of various kinds such as convex programming, discrete optimization, nonlinear programming, etc. In more complex situations, the objective functions that determine the effectiveness of decisions can be multiextremal, and decision making will require solving global optimization problems. In the most general case, however, the effectiveness of decisions may be determined by several objective criteria, which makes it necessary to solve multi-objective optimization (MOO) problems. Thus, MOO problems are the most common decision making statements that one has to solve in many scientific and technical applications. A large number of approaches and methods have been developed to solve such problems and they have been used to solve many decision making problems in various fields of practical applications -- see e.g. \cite{c1,c2,c3,c4,c5,c6,c7,c8,c9}.

@@

It is also important to note that by solving simultaneously a large number of scalar optimization problems thus generated it is possible to significantly decrease computational complexity of each separate problem. This effect is achieved due to the fact that all such scalar problems  are based on the same MOO problem and, consequently, all computed values of the efficiency criteria of the MOO problem can be reduced to values of any scalar problem being solved simultaneously without any time consuming calculations. In such cases, all the search information obtained in solving any single scalar problem can be used for solving all other scalar problems of the same set.

This paper presents the results of the authors' continuing research aimed to develop efficient methods for solving MOO problems \cite{c35,c36,c37,c44}. In \cite{c35}, some methods for solving global optimization problems are considered, in \cite{c36},  a general decision-making model based on MOO problems is proposed, and in \cite{c37}, a theoretical analysis is presented regarding the effectiveness of the use of search information when solving MOO problems. In \cite{c44}, we consider MOO problems in which the efficiency criteria can be ordered by importance (lexicographic multi-objective optimization). In this case, MOO problems are reduced to global optimization problems with non-convex constraints for which constrained optimization methods should be applied. In this paper, we propose new computational schemes for using search information that significantly reduce the amount of calculations required to solve MOO problems.

@@

The reference decision $y^0 \in D$ in (\ref{eq:05}) can be known \textit{a priori} or determined on the basis of some known prototype. In many cases, an abstract ideal decision $y^0 \in D$ is used as the reference decision $y^*\in D$ in which the objective functions $f_i$, $1 \leq i \leq s$, have minimal possible values, i.e.
\begin{equation}
\label{eq:07}
f_i^* =f_i (y^*)=\min_{y \in D} f_i (y), 1 \leq i \leq s.
\end{equation}

@@

Using the dimensionality reduction results in one more additional advantage of the proposed approach: many well-known one-dimensional global search algorithms (possibly, after some additional generalisation) can be used to solve the initial multidimensional MOO problem from (\ref{eq:01}) \cite{c22,c23,c24,c25,c26,c27,c28,c29}. At the same time, it should be noted that most works where dimensionality reduction is a key feature for solving multiextremal problems like (\ref{eq:04}) rely on the information-statistical theory of global search \cite{c15}. This theory has provided the basis for developing a large number of efficient methods for multiextremal optimization \cite{c28,c29,c30,c31,c32,c33,c34,c35,c36,c37}.

@@

Convergence conditions of the algorithms are formulated in the form (\ref{eq:16})--(\ref{eq:18}) depend on the properties of the interval characteristics used. One of sufficient conditions for convergence of algorithms is, for example, the requirement that the characteristic of the interval containing the global minimum point should take on the maximum value at step 1 of the scheme (\ref{eq:16})--(\ref{eq:18}) during the global search iterations. This condition is satisfied, for example, for the multidimensional generalized algorithms proposed in \cite{c22,c23} when the H{\"o}lder constant from (\ref{eq:10}) is specified exactly. For GSA, a sufficient condition for convergence is the relation 
\begin{equation}
\label{eq:20}
m \geq 2^{3-1⁄N} L\sqrt{N+3},
\end{equation}
which must be fulfilled starting from some iteration $k>1$ of the global search ($L$ is the Lipschitz constant from (\ref{eq:08})) \cite{c15}. Moreover, if condition (\ref{eq:20}) is satisfied, only the points of the global minimum of the function $\varphi(y)$ from (\ref{eq:04}) will be the limit points of the trial sequence $\{y^k=y(x^k)\}$ generated by the GSA algorithm.

@@

As mentioned earlier, in the process of solving the MOO problem, it may be necessary to find several different efficient decisions due to possible changes in the optimality requirements. Obtaining different efficient decisions in the proposed approach is achieved by choosing different convolution coefficients (importance indicators) for the objective functions $f_i(y)$, $1 \leq i \leq s$, which results in obtaining different  scalar multiextremal functions $F(\lambda, y)$ from (\ref{eq:04}). These functions $F(\lambda, y)$ can be optimized sequentially. This determines a multi-stage scheme for solving the MOO problem, when at each stage DM specifies the necessary scalarization coefficients, then the resulting scalar optimization problem is solved, after which DM analyzes the efficient decisions found.

@@

As shown in Section \ref{sec:03}, the information-statistical multiextremal optimization algorithms used in the proposed approach determine the points of consecutive iterations of the global search taking into account the search information
\begin{equation}
\label{eq:22}
A_k=\{(x_i,z_i,f_i )^T:1 \leq i \leq k\}
\end{equation}
obtained in the calculations (see (\ref{eq:14})--(\ref{eq:17})). In (\ref{eq:22}) $x_i$, $1 \leq i \leq k$,  are the reduced points of performed global search iterations ordered in ascending order of coordinates,  $z_i$, $f_i$, $1 \leq i \leq k$, are the values of the scalar function $F(\lambda, y)$ from (\ref{eq:04}) and the objective functions $f_i(y)$, $1 \leq i \leq s$, from (\ref{eq:01}) of the current optimization problem to be solved at the points $x_i=y(x_i)$, $1 \leq i \leq k$. By using search information $A_k$ from (\ref{eq:22}) when choosing next search iterations it is possible to solve global optimization problems more efficiently and to provide convergence of algorithms only to the global minima of multiextremal functions being minimized.

@@

For MGSA, 100 problems (\ref{eq:04}) with different values of convolution coefficients $\lambda$ uniformly distributed in $\Lambda$ were solved. The results of the experiments performed are presented in Table \ref{tab:1}.

@@

The results of numerical experiments are presented in Table \ref{tab:2} and Fig. \ref{fig:1}. The Table shows in the first column the number of functions $\Phi_p (y)$ from (\ref{eq:21}) being optimized.  The columns named Iters contain the average number of iterations performed by the algorithm to solve the MOO problem. The columns named HV and DU contain the values of the HV and DU indicators. The results of the numerical experiments are divided with respect to the three algorithms used: GSA is described in Section \ref{sec:03}, MGSA is discussed in Subsection \ref{subsec:32}, SGSA is presented in Subsection \ref{subsec:43}.

@@

The experimental results show that by reusing search information when optimizing 25 functions from the set $\Phi_p (y)$, the total computational iterations can be reduced  by more than 9.5 times without resorting to any additional computational resources. At the same time, the MGSA algorithm performs the least number of iterations (the number of calculations of objective function values), while the SGSA algorithm provides a better approximation of the Pareto domain both in terms of approximation completeness and coverage uniformity. It is also noteworthy that the average number of iterations to optimize one function from the set $\Phi_p (y)$ decreases by more than 4 times as the number of functions in the set $\Phi_p (y)$ increases -- see Table \ref{tab:3}.

@@

In the third series of numerical experiments, 10 bi-criteria five-dimensional MOO problems were solved, i.e. $N = 5$, $s = 2$. The objective functions of the MOO problems were determined, as before, using the GKLS generator \cite{c43}. When solving the problems, the accuracy $\varepsilon=0.05$ and the reliability $r=5.6$ were used. The results of numerical experiments are presented in Table \ref{tab:4} and Fig. \ref{fig:2}.
